Approximation as Differentiation

### 2025-07-08 main/approx-as-tangents.tex
[ tex ] talk about chain rule more

\bob{mention chain rule here? and tangent spaces?}

@@

In a partially ordered set $X$, for any $x \in X$ the set of elements below $x$, $\downset{x} = \{ x' \mid x' \sqsubseteq x \}$, is itself a partially ordered set. These approximations of $x$ we will think of as ``tangents'' at $x$, and the whole set $\downset{x}$ as the ``tangent space''\footnote{\bob{Is it possible to derive the tangent spaces by thinking about continuous functions passing through $x$?}}. Tangent spaces are vector spaces, so in the partially ordered setting we take elements of $\downset{x}$ to be approximations of processes defined at $x$. As we can add tangents, we assume we can take meets of approximations in $\downset{x}$:

@@

\begin{theorem}
  Bounded meet lattices and conditionally multiplicatve functions form
  a category $\CM$. This category has products, coproducts and
  exponentials.
\end{theorem}

\begin{proof}
  \AGDA. See \citet{amadio-curien} for the case when the posets are
  also cpos. The crucial technical step, identified by
  \citet{berry79}, is that the ordering on conditionally
  multiplicative functions is not the extensional ordering
  ($f \sqsubseteq g$ iff $\forall x. f(x) \subseteq g(x)$) but instead
  the stable ordering $f \sqsubseteq g$ iff \bob{FIXME}.
\end{proof}

@@

Given these examples, conditional multiplicativity seems to be a reasonable analogue to functions with a well-defined notion of derivative. However, for \GPS we also require an analogue to the reverse derivative, where we map approximations backwards to give the least approximation of the input for a given approximation of the output. In the case of smooth functions, we are always guaranteed a reverse derivative but there is not always a best way to map approximations backwards for cm functions, as the following example shows.

\begin{example}[Is Conditional Multiplicativity Enough?]
  \label{ex:non-stable-function}
  % In the preceeding examples, we demonstrated conditional
  % multiplicativity by giving a backwards mapping of approximations
  % that forces preservation of meets. A natural question is whether or
  % not we always have such a backwards map which we could use for
  % backwards propagation of approximations.
  An example that is conditionally multiplicative, but does not admit
  a backards map of approximations (from \citet{amadio-curien} just
  before Lemma 12.2.3, originally due to Berry) is given by defining
  $\mathrm{unstable} : D \to \{\bot, \top\}$, where
  $D = \bot \sqsubseteq \cdots \sqsubseteq n \sqsubseteq \cdots
  \sqsubseteq 1 \sqsubseteq 0$, as $\mathrm{unstable}(\bot) = \bot$
  and $\mathrm{unstable}(n) = \top$. This is monotone, and preserves
  meets in every $\downset{x}$. However, there is no ``best'' (i.e.,
  least) input that gives us any finite output.
\end{example}

@@

In light of the \exref{non-stable-function}, we consider
\citet{berry79}'s definition of \emph{stable function} that requires
the existence of a reverse mapping directly, even without assuming
that any meets exist:

@@

\begin{theorem}
  L-posets and stable functions form a category with products and
  coproducts.
\end{theorem}

@@

Our goal is to model a higher-order language suitable for writing
queries on databases as in \exref{introduction-example}, so why should
we not just take complete L-posets and stable functions as our model?
We have two reasons for moving to a slightly different model in \secref{model}:

\begin{enumerate}
\item Even without completeness, in Bounded meet posets and L-posets,
  values and their approximations live in the same set. However, in
  the \emph{total} query language we wish to model in
  \secref{language}, we are not directly interested in the behaviour
  of programs on approximations as we would be for partial programs
  with general recursion. (Moreover, in the light of
  \exref{intervals-and-maxima-elements} it is not clear whether
  approximations for partiality and approximations for stability ought
  to be the same thing. \bob{fwd reference to discussion.}) In
  \exref{intervals-and-maxima-elements}, maximal elements could be
  taken to be the ``proper values''. One idea is to restrict to
  conditionally multiplicative or stable functions that preserve
  maximal elements. However, this idea fails at higher order:
  functions that take maximal elements to maximal elements are not
  themselves maximal elements. We could devise a category of L-posets
  with totality predicates (which would pick out maximal elements at
  first-order) and totality preserving functions, but we prefer a more
  direct method of separating values proper from their approximations
  using the \emph{Category of Families} construction we detail in the
  next section.
\item A more local reason is that we wish to formalise our
  construction in the proof assistant Agda \cite{agda} in order to get
  an \emph{executable} model of the language in
  \secref{language}. Agda's type theory is both predicative and
  constructive. Predicativity means that it does not have complete
  lattices in the classical sense: for a type $X$ we can only get
  suprema and infinma of families in a lower universe level than
  $X$. This is not necessarily a problem, as \citet{dejong21} shows
  how to develop a large amount of domain theory in a predicative
  setting. However, due to constructivity the lifting construction
  (due to \citet{escardo-knapp}) requires a proof of definedness to
  extract a value. Since we are modelling a total language, we prefer
  to have a model that computes directly.
\end{enumerate}

@@

We have seen that \citet{berry79}'s theory of stable functions between suitable partial orders can be seen as form of differentiability with forward and reverse derivatives. In the next section, we describe a model based on these ideas suitable for constructing executable models of total higher order languages.

@@

As well as codifying exactly what we mean by differentiable structure
on partially ordered sets, proving this conjecture would also tell us
what higher derivatives mean in this context as well, something that
we have not considered above. We will extend this conjecture to our
refined model of partially ordered approximations in
\secref{model-summary}.

@@

% \bob{Put a conjecture about Tangent Categories in somewhere}

% Given this close connection between \GPS and differentiable programming, we can take structures intended for modelling automatic differentiation and use them to model \GPS. This will enable us to generalise and expand the scope of \GPS to act as a foundation for data provenance in a wide range of computational settings.

>>>>

\begin{remark}[Chain Rule]
  \label{rem:chain-rule}
  A useful property of derivative maps is that they compose according
  to the chain rule. Suppose that $f : M \to N$ and $g : N \to K$ are
  smooth functions. Then for any $x \in M$, we have:
  \begin{itemize}
  \item $\pushf{(g \circ f)}_x = \pushf{g}_{f(x)} \circ \pushf{f}_x : \tangents_x(M) \linearto \tangents_{g(f(x)}(K)$
  \item $\pullf{(g \circ f)}_x = \pullf{f}_x \circ \pullf{g}_{f(x)} : \cotangents_{f(x)}(K) \linearto \cotangents_x(M)$
  \end{itemize}
  The chain rule has the practical effect that we can compute
  derivative maps of $f$ and $g$ independently and compose them,
  instead of the potentially more difficult task of computing the
  derivative maps of $g \circ f$. As we shall see below, stable maps
  also obey a chain rule, and this forms the basis of the general
  categorical approach to differentiability that we describe in
  \secref{models-of-total-gps}.
\end{remark}

@@

In a partially ordered set $X$, for any $x \in X$ the set of elements below $x$, $\downset{x} = \{ x' \mid x' \sqsubseteq x \}$, is itself a partially ordered set. These approximations of $x$ we will think of as ``tangents'' at $x$, and the whole set $\downset{x}$ as the ``tangent space''. Tangent spaces are vector spaces, so in the partially ordered setting we take elements of $\downset{x}$ to be approximations of processes defined at $x$. As we can add tangents, we assume we can take meets of approximations in $\downset{x}$:

% \footnote{\bob{Is it possible to derive the tangent spaces by
% thinking about continuous functions passing through $x$?}}

@@

\begin{theorem}
  Bounded meet lattices and conditionally multiplicatve functions form
  a category $\CM$. This category has products, coproducts, and
  exponentials.
\end{theorem}

\begin{proof}
  \AGDA. See \citet{amadio-curien} (FIXME: theorem reference) for the
  case when the posets are also cpos. The crucial technical step,
  identified by \citet{berry79}, is that the ordering on conditionally
  multiplicative functions is not the extensional ordering
  ($f \sqsubseteq_{\mathrm{ext}} g$ iff
  $\forall x. f(x) \sqsubseteq g(x)$) but instead the stable ordering:
  $f \sqsubseteq_{\mathrm{st}} g$ iff $f \sqsubseteq_{ext} g$ and
  $\forall x, x'.\,x \sqsubseteq x' \Rightarrow f(x) = f(x') \land
  g(x)$.
\end{proof}

\begin{remark}[Chain Rule]
  \label{rem:cm-chain-rule}
  It is almost a triviality at this point, but the crucial point is
  that, for any cm functions $f : X \to Y$ and $g : Y \to Z$, the
  restriction maps (``forward derivatives'') compose according to the
  chain rule from \remref{chain-rule}:
  \begin{displaymath}
    (g \circ f)_x = g_{f(x)} \circ f_x
  \end{displaymath}
  We will see this phenomenon repeated in the definition of stable
  functions below.
\end{remark}

@@

Given these examples, conditional multiplicativity seems to be a
reasonable analogue to functions with a well-defined notion of
derivative. For \GPS we also require an analogue to the reverse
derivative, where we map approximations backwards to give the least
approximation of the input for a given approximation of the output. In
the case of smooth functions, we are always guaranteed a reverse
derivative. However, there is not always a best way to map
approximations backwards for cm functions, as the following example
shows.

\begin{example}[Is Conditional Multiplicativity Enough?]
  \label{ex:non-stable-function}
  An example that is conditionally multiplicative, but does not admit
  a backards map of approximations (from \citet{amadio-curien} just
  before Lemma 12.2.3, originally due to Berry) is given by defining
  $\mathrm{unstable} : D \to \{\bot \sqsubseteq \top\}$, where
  $D = \bot \sqsubseteq \cdots \sqsubseteq n \sqsubseteq \cdots
  \sqsubseteq 1 \sqsubseteq 0$, as $\mathrm{unstable}(\bot) = \bot$
  and $\mathrm{unstable}(n) = \top$. This is monotone, and preserves
  meets in every $\downset{x}$. However, there is no ``best'' (i.e.,
  least) input that gives us any finite output.
\end{example}

@@

In light of the \exref{non-stable-function}, we turn to
\citet{berry78}'s definition of \emph{stable function} that requires
the existence of a reverse mapping directly, even without assuming
that any meets exist:

@@

\begin{theorem}
  L-posets and stable functions form a category $\Lposet$ with products
  and coproducts.
\end{theorem}

\begin{remark}[Chain Rule]
  As for the ``forward derivatives'' of conditionally multiplicative
  functions, the forward and backwards parts of a stable function
  compose according to the chain rule (c.f. \remref{chain-rule}):
  \begin{itemize}
  \item $(g \circ f)_x = g_{f(x)} \circ f_x : \downset{x} \linearto \downset{g(f(x))}$
  \item $(g \circ f)^*_x = f*_x \circ g^*_{f(x)} : \downset{g(f(x))} \linearto \downset{x}$
  \end{itemize}
  We will use this property in \propref{embed-stable} to show that
  $\Lposet$ embeds into our category of sets-with-approximation.
\end{remark}

@@

Our goal is to model a higher-order language suitable for writing
queries on databases as in \exref{introduction-example}, so why should
we not just take complete L-posets and stable functions as our model
of \GPS?  We have two reasons for moving to a different model in
\secref{models-of-total-gps}:

\begin{enumerate}
\item Even without completeness, in Bounded meet posets and L-posets,
  values and their approximations live in the same set. However, in
  the \emph{total} query language we wish to model in
  \secref{language}, we are not directly interested in the behaviour
  of programs on approximations as we would be for partial programs
  with general recursion. (Moreover, in the light of
  \exref{intervals-and-maxima-elements} it is not clear whether
  approximations for partiality and approximations for stability ought
  to be the same thing. \bob{fwd reference to discussion.}) In
  \exref{intervals-and-maxima-elements}, maximal elements could be
  taken to be the ``proper values''. One idea is to restrict to
  conditionally multiplicative or stable functions that preserve
  maximal elements. However, this idea fails at higher order:
  functions that take maximal elements to maximal elements are not
  themselves maximal elements. We could devise a category of L-posets
  with totality predicates (which would pick out maximal elements at
  first-order) and totality preserving functions, but we prefer a more
  direct method of separating values proper from their approximations
  using the \emph{Category of Families} construction we detail in the
  next section.
\item A more practical reason is that we wish to formalise our
  construction in the proof assistant Agda \cite{agda} in order to get
  an \emph{executable} model of the language in
  \secref{language}. Agda's type theory is both predicative and
  constructive. Predicativity means that it does not have complete
  lattices in the classical sense: for a type $X$ we can only get
  suprema and infinma of families in a lower universe level than
  $X$. This is not necessarily a problem, as \citet{dejong21} shows
  how to develop a large amount of domain theory in a predicative
  setting. However, due to constructivity the lifting construction
  (due to \citet{escardo-knapp}) requires a proof of definedness to
  extract a value. Since we are modelling a total language, we prefer
  to have a model that computes directly.
\end{enumerate}

@@

We have seen that \citet{berry78}'s theory of stable functions between suitable partial orders can be seen as form of differentiability with forward and reverse derivatives. In the next section, we describe a model based on these ideas suitable for constructing executable models of total higher order languages.

@@

As well as codifying exactly what we mean by differentiable structure
on partially ordered sets, proving this conjecture would also tell us
what higher derivatives mean in this context as well, something that
we have not considered above. We will extend this conjecture to our
refined model of lattice approximated sets in \secref{model-summary}.

@@
